---
title: "Flow_CATALYST"
author: "docaspar"
date: "12/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, root.dir = ".")
```



# Libraries
```{r libs, message=FALSE}
library(tidyverse)
library(SingleCellExperiment)
library(CATALYST)
library(flowCore)
library(xlsx)
library(readxl)
library(clustree)
library(diffcyt)
library(Rphenograph)
library(edgeR)
library(MetBrewer)
library(ggpubr)
library(pals)
library(ComplexHeatmap)
```


# Set paths
```{r set paths, warning=FALSE}
rm(list = ls())

path.data <- "../Data/20220505_Live"
path.files <- "../Files/"
path.out <- "../R_out/Unsupervised_10k_final/"

colvec <- met.brewer("Monet",n=20)[c(4,12,10,18,19)]
```


# Create Meta
This chunk creates the meta data from the file names (list.files())
```{r create meta, warning=FALSE}
#create meta
file.names <- list.files(path.data, pattern = "fcs$")
md <- data.frame("file_name" = file.names)
md <- md %>%
  mutate(copy = file_name) %>%
  separate(copy, c("export", "well", "patient_id", "condition", "live", "fcs")) %>%
  dplyr::filter(patient_id != "FMO") %>%
  dplyr::select(-c("export", "well", "live", "fcs"))

md[!md$condition %in% c("V2", "V3", "V8", "V9"), "condition"] <- "H"
md <- md %>% mutate("sample_id" = paste(patient_id, condition, sep = "_"))
md$condition <- as.factor(md$condition)
md <- md[order(md$condition, md$patient_id),]
md$sample_id <- factor(md$sample_id, levels = md$sample_id)
md %>% head()
```


# Create Single Cell Experiment (SCE) object
Here, we read the .fcs files as a flow set and downsample to 10000 cells per sample.
Next, we convert the flow set to an SCE object (prepData()) and transform with cofactor 6000.
To save time, I saved the resulting SCE object. It is imported again in the next chunk.
```{r create flowset}
fs <- read.flowSet(md$file_name, path = path.data, transformation = FALSE, truncate_max_range = FALSE)

fsApply(fs, nrow)
n = 10000
dsFilt <- sampleFilter(size = n, filterId="dsFilter")
result <- flowCore::filter(fs, dsFilt)

for(i in 1:length(result)){
  a <- result@.Data[[i]]@subSet
  l <- a %>% length()
  if(n>l){
    next
  }
  set.seed(42)
  ids <- sample(1:l, n, replace=FALSE) %>% sort()
  x <- rep(FALSE, l)
  x[ids] <- TRUE
  result@.Data[[i]]@subSet <- x
}

fs.ds <- Subset(fs, result) 
fsApply(fs.ds, nrow) %>% head()
remove(fs, a, dsFilt)

panel <- as.data.frame(markernames(fs.ds))
colnames(panel) <- "antigen"
panel$channel <- rownames(panel)
rownames(panel) <- panel$marker_name
panel$marker_class <- "state"

sce <- prepData(x = fs.ds, md = md, panel = panel, panel_cols = list(channel = "channel",
        antigen = "antigen"), transform = TRUE, cofactor = 6000, FACS = TRUE)

#Add metadata to SCE
colData(sce)$cell_id <- 1:nrow(colData(sce))
rowData(sce)$channel_id <- 1:nrow(rowData(sce))

levels(sce$sample_id)
```



# Save SCE
```{r}
#Save SCE object
saveRDS(sce, file = paste0(path.files, "SCE_downsammpled_10k.rds"))
```


# Session Info
```{r}
sessionInfo()
```

