---
title: "Flow_CATALYST"
author: "docaspar"
date: "12/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, root.dir = ".")
```



# Libraries
```{r libs, message=FALSE}
library(tidyverse)
library(SingleCellExperiment)
library(CATALYST)
library(rstudioapi)
library(flowCore)
#library(xlsx)
library(readxl)
library(clustree)
library(diffcyt)
library(Rphenograph)
library(edgeR)
library(MetBrewer)
library(ggpubr)
library(pals)
library(ComplexHeatmap)
library(cowplot)
library(rasterpdf)
```


# Set paths
```{r set paths, warning=FALSE}
rm(list = ls())

path.data <- "../Data/20220505_Live"
path.files <- "../Files/"
path.out <- "../R_out/Unsupervised_10k_final/"

colvec <- met.brewer("Monet",n=20)[c(4,12,10,18,19)]

source("flow_functions.R")
```

# Import SCE object

Import SCE object that we saved in the previous chunk.
```{r load and coldata}
sce <- readRDS(file = paste0(path.files, "SCE_downsammpled_10k.rds"))
sce
ei(sce) %>% head()
colData(sce)
```

# Add marker_classes to rowdata
```{r rowdata}
primary_markers <- c("CD3", "CD19", "CD14", "CD16", "CD56", "CD4", "CD8", "HLADR", "CD25")
rowData(sce)[rowData(sce)$marker_name %in% primary_markers, "marker_class"] <- "type"
rowData(sce)[!rowData(sce)$marker_name %in% primary_markers, "marker_class"] <- "state"

type_markers(sce)
state_markers(sce)
```

# Exploratory Data Analysis
```{r Health check}
#pdf(paste0(path.out, "0_Data_exploration.pdf"), onefile = TRUE, height = 9, width = 16)

plotCounts(sce,
    group_by = "sample_id",
    color_by = "condition")

# plotScatter(sce, c("CD3", "CD19"))
# plotScatter(sce, c("CD4", "CD8"))
# plotScatter(sce, c("CD14", "CD16"))

set.seed(1234)
sce <- runDR(sce, dr = "UMAP", features = "type", cells = 500)

plotDR(sce, "UMAP", color_by = "patient_id")
plotDR(sce, "UMAP", color_by = c(type_markers(sce), state_markers(sce)), ncol = round(sqrt(16*length(c(type_markers(sce), state_markers(sce)))/9)))
#dev.off()
```

# Primary 
## flowsom clustering
Here, we perform the primary clustering and dimensionality reduction (UMAP).
```{r primary clustering}
type_markers(sce) #will be used for clustering
state_markers(sce) # will not be used for clustering

set.seed(1234)
kinit <- 30 #try 10 or 20
sce <- cluster(sce, features = type_markers(sce),
    xdim = 10, ydim = 10, maxK = kinit)

#pdf(paste0(path.out, "1_Primary_clustering.pdf"), onefile = TRUE, height = 9, width = 16)

clustree(cluster_codes(sce), prefix = "meta")

plotExprHeatmap(sce, features = "type",
    by = "cluster_id", k = paste0("meta", kinit),
    scale = "first", q = 0.01, perc = TRUE, col_dend = FALSE, bars = TRUE, col_clust = FALSE)

#plotClusterExprs(sce, k = paste0("meta", kinit), features = "type") + labs(title = "primary clustering marker expression histogram")

plotDR(sce, "UMAP", color_by = paste0("meta", kinit)) +
  theme_cowplot()
plotDR(sce, "UMAP", color_by = type_markers(sce), ncol = round(sqrt(16*length(type_markers(sce))/9)), assay = "exprs")+
  theme_cowplot()+ 
  theme(strip.background = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())
  
#dev.off()

code <- cluster_codes(sce)[,c("som100", paste0("meta", kinit))]
coldata <- as.data.frame(colData(sce))
coldata <- merge(coldata, code, by.x = "cluster_id", by.y = "som100")
colData(sce) <- DataFrame(coldata[order(coldata$cell_id),])
```



## Merge primary clustering
Label and merge clusters. 
```{r merge primary clusters}
(merging_table <- read_excel(paste0(path.files, "merged_clusters_primary2.xlsx")))
                            
sce <- mergeClusters(sce, k = paste0("meta", kinit), table = merging_table, id = "primary", overwrite = TRUE)

code <- cluster_codes(sce)[,c("som100", "primary")]
coldata <- as.data.frame(colData(sce))
coldata <- merge(coldata, code, by.x = "cluster_id", by.y = "som100")
colData(sce) <- DataFrame(coldata[order(coldata$cell_id),])

#pdf(paste0(path.out, "2_Primary_merging.pdf"), onefile = TRUE, height = 9, width = 16)
plotDR(sce, "UMAP", color_by = "primary") +
  scale_color_brewer(palette = "Paired")+
  theme_cowplot()+
  theme(axis.ticks = element_blank(), 
        axis.text = element_blank())

plotDR(sce, "UMAP", color_by = type_markers(sce), ncol = round(sqrt(16*length(type_markers(sce))/9)))+
  theme_cowplot()+
  theme(strip.background = element_blank(),
        axis.ticks = element_blank(), 
        axis.text = element_blank())
#dev.off()

raster_pdf(filename = paste0(path.out, "2a_Primary_merging.pdf"), height = 9, width = 11, res = 200)
plotDR(sce, "UMAP", color_by = "primary") +
  scale_color_brewer(palette = "Paired")+
  theme_cowplot()+
  theme(axis.ticks = element_blank(), 
        axis.text = element_blank())

plot_DR(x = sce, color_by = "primary", add_labels = TRUE)+
  theme_cowplot()
dev.off()

#raster_pdf(filename = paste0(path.out, "2b_Primary_merging.pdf"), height = 5, width = 11, res = 200)
plotDR(sce, "UMAP", color_by = type_markers(sce), ncol = 5)+
  theme_cowplot()+
  theme(strip.background = element_blank(),
        axis.ticks = element_blank(), 
        axis.text = element_blank())
#dev.off()

#raster_pdf(filename = paste0(path.out, "2b_Primary_merging_long.pdf"), height = 9, width = 10, res = 200)
plotDR(sce, "UMAP", color_by = type_markers(sce), ncol = 3)+
  theme_cowplot()+
  theme(strip.background = element_blank(),
        axis.ticks = element_blank(), 
        axis.text = element_blank())
#dev.off()

```


# Secondary
## filter, change markers & run UMAP
```{r filter for secondary clustering}
scecd4 <- sce[,colData(sce)$primary == "CD4 T cells"]

cd4_markers <- c("CD127", "CD25", "Foxp3")
rowData(scecd4)[rowData(scecd4)$marker_name %in% cd4_markers, "marker_class"] <- "type"
rowData(scecd4)[!rowData(scecd4)$marker_name %in% cd4_markers, "marker_class"] <- "state"

type_markers(scecd4)

set.seed(1234)
scecd4 <- runDR(scecd4, dr = "UMAP", features = "type", cells = 500)
plotDR(scecd4, "UMAP", color_by = type_markers(scecd4), ncol = 3, assay = "exprs")+
  theme_cowplot()+
  theme(strip.background = element_blank(),
        axis.ticks = element_blank(), 
        axis.text = element_blank())

```


## Secondary clustering
```{r secondary clustering}
set.seed(1234)
kinit <- 15 #try 10 or 20
scecd4 <- cluster(scecd4, features = type_markers(scecd4),
    xdim = 10, ydim = 10, maxK = kinit)

#pdf(paste0(path.out, "3_Secondary_clustering.pdf"), onefile = TRUE, height = 9, width = 16)

clustree(cluster_codes(scecd4), prefix = "meta")

plotExprHeatmap(scecd4, features = "type",
    by = "cluster_id", k = paste0("meta", kinit),
    scale = "first", q = 0.01, perc = TRUE, col_dend = FALSE, bars = TRUE, col_clust = FALSE)

#plotClusterExprs(scecd4, k = paste0("meta", kinit), features = "type") + labs(title = "primary clustering marker expression histogram")

plotDR(scecd4, "UMAP", color_by = paste0("meta", kinit))+
  theme_cowplot()+
  theme(axis.ticks = element_blank(), 
        axis.text = element_blank())
plotDR(scecd4, "UMAP", color_by = type_markers(scecd4), ncol = 3, assay = "exprs")+
  theme_cowplot()+
  theme(strip.background = element_blank(),
        axis.ticks = element_blank(), 
        axis.text = element_blank())
#dev.off()


# time.start <- Sys.time()
# 
# set.seed(1234)
# phenoresults <- Rphenograph(t(assay(filterSCE(scecd4, marker_class == "type"), "exprs")), k = 200)
# 
# #Add phenograph cluster_ids to scecd4 object
# colData(scecd4)$pheno <- as.factor(membership(phenoresults[[2]]))
# scecd4$cluster_id <- as.factor(membership(phenoresults[[2]]))
# metadata(scecd4)$cluster_codes <- data.frame(
#     custom = factor(levels(as.factor(membership(phenoresults[[2]]))), levels = levels(as.factor(membership(phenoresults[[2]])))))
# 
# pdf(paste0(path.out, "3_Secondar_clustering_k100 .pdf"), onefile = TRUE, height = 9, width = 16)
# plotExprHeatmap(scecd4, features = "type",
#     by = "cluster_id", k = "custom",
#     scale = "first", q = 0.01, perc = TRUE, col_dend = FALSE, bars = TRUE, col_clust = FALSE)
# 
# plotDR(scecd4, "UMAP", color_by = "custom")
# plotDR(scecd4, "UMAP", color_by = type_markers(scecd4), ncol = 3, assay = "exprs")
# dev.off()
# 
# time.finish <- Sys.time()
# time.finish - time.start
```


## Cluster merging
```{r merge secondary cluster}
(merging_table_secondary <- read_excel(paste0(path.files, "merged_clusters_cd4_v4.xlsx")))

scecd4 <- mergeClusters(scecd4, k = "meta15", table = merging_table_secondary, id = "secondary")

code <- cluster_codes(scecd4)[,c("som100", "meta15", "secondary")]
coldata <- as.data.frame(colData(scecd4))
coldata <- merge(coldata, code, by.x = "cluster_id", by.y = "som100")
colData(scecd4) <- DataFrame(coldata[order(coldata$cell_id),])

#pdf(paste0(path.out, "04_Secondary_merging.pdf"), onefile = TRUE, height = 9, width = 16)
plotDR(scecd4, "UMAP", color_by = "secondary") +scale_color_brewer(palette = "Paired")+
  theme_cowplot()+
  theme(axis.ticks = element_blank(), 
        axis.text = element_blank())
plotDR(scecd4, "UMAP", color_by = type_markers(scecd4), ncol = 3)+
  theme_cowplot()+
  theme(strip.background = element_blank(),
        axis.ticks = element_blank(), 
        axis.text = element_blank())
#dev.off()

#raster_pdf(filename = paste0(path.out, "04a_Secondary_merging.pdf"), height = 9, width = 11, res = 200)
plotDR(scecd4, "UMAP", color_by = "secondary") +scale_color_brewer(palette = "Paired")+
  theme_cowplot()+
  theme(axis.ticks = element_blank(), 
        axis.text = element_blank())
#dev.off()

#raster_pdf(filename = paste0(path.out, "04b_Secondary_merging.pdf"), height = 9, width = 11, res = 200)
plotDR(scecd4, "UMAP", color_by = type_markers(scecd4), ncol = 2)+
  theme_cowplot()+
  theme(strip.background = element_blank(),
        axis.ticks = element_blank(), 
        axis.text = element_blank())
#dev.off()
```

# Feed back subclustering
Feed back subclustering into main SCE object
```{r feed back subclustering}
coldata.sce <- colData(sce)
coldata.cd4 <- colData(scecd4)
coldata <- merge(coldata.sce, coldata.cd4[,c("cell_id", "secondary")], by = "cell_id", all = TRUE)
coldata$primary <- as.character(coldata$primary)
coldata$secondary <- as.character(coldata$secondary)
coldata[is.na(coldata$secondary), "secondary"] <- coldata[is.na(coldata$secondary), "primary"]
coldata$secondary <- as.factor(coldata$secondary)

colData(sce) <- DataFrame(coldata[order(coldata$cell_id),])

#pdf(paste0(path.out, "5_Feeding_back_subclusters.pdf"), onefile = TRUE, height = 9, width = 16)
plotDR(sce, "UMAP", color_by = "secondary") +scale_color_manual(values = unname(cols25()))+
  theme_cowplot()+
  theme(axis.ticks = element_blank(), 
        axis.text = element_blank())
plotDR(sce, "UMAP", color_by = type_markers(sce), ncol = round(sqrt(16*length(type_markers(sce))/9)))+
  theme_cowplot()+
  theme(strip.background = element_blank(),
        axis.ticks = element_blank(), 
        axis.text = element_blank())
#dev.off()
```


## DA visualisation
```{r DA}
dtp <- colData(sce)[c("secondary", "sample_id")] %>% 
  table() %>% 
  as.data.frame() %>%
  mutate("copy" = sample_id) %>%
  separate(copy, c("patient_id", "condition")) %>%
  dplyr::rename("counts" = Freq)

totdtp <- colData(sce)[c("sample_id")] %>% table() %>% data.frame() %>% dplyr::rename("sample_id" = Var1, "totcounts" = Freq) #counts per sample to normalize counts by sample size. irrelevant in this case, since all samples have the same samplesize.
dtp <- merge(dtp, totdtp) %>% mutate("Freq" = counts/totcounts*100)


#pdf(paste0(path.out, "6_Cluster_Abundance.pdf"), height = 9, width = 16)

comparisons <- list(c("V2", "V3"), c("V8", "V9"), c("V2", "V9"))
ggplot(dtp, aes(x = condition, y = Freq, col = condition))+
  geom_violin(aes(fill = condition), alpha = 0.5, draw_quantiles = 0.5)+
  #geom_jitter(width = 0.1)+
  geom_point()+
  geom_line(aes(group = patient_id), col = "grey")+
  facet_wrap("secondary", scales = "free_y")+
  theme_bw()+
  labs(col = "Time point", fill = "Time point", title = "Cluster Abundance per Time point", y = "Frequency of PBMCs (%)")+
  theme(axis.title.x = element_blank(),
        plot.title = element_text(hjust = 0.5),
        strip.background = element_blank())+
  scale_color_manual(values = colvec)+
  scale_fill_manual(values = colvec)+
  stat_compare_means(comparisons = comparisons, method = "t.test", paired = TRUE)+
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))


#dev.off()

dtp[dtp$secondary == "CD19 B cells" & dtp$condition %in% c("V2", "V9"), c("patient_id", "condition", "Freq")]
```

# Stacked bar plot
Create stacked bar plot from cell type frequencies.
```{r, fig.width=16}
cur_df <- dtp %>%
  dplyr::filter(condition != "H") %>%
  dplyr::select(sample_id, secondary, Freq) %>%
  spread(key = secondary, value = Freq)

matrixrownames <- cur_df$sample_id 

# now we create a matrix from the data and cluster the data based on the cell fractions
hm_dat = as.matrix(cur_df[,-1])
rownames(hm_dat) <- as.character(matrixrownames)

# calculate distance and then cluster samples on cluster fraction
dd <- dist((hm_dat))
hc <- hclust(dd, method = "ward.D2")
row_sorted <- hc$labels

# now we generate the clinical metadata and order it in the same way as the celltype data
mmd <- cur_df %>% 
  dplyr::select(sample_id) %>%
  dplyr::mutate("copy" = sample_id) %>%
  separate(copy, into = c("patient_id", "timepoint"), sep = "_") %>%
  as.matrix() %>%
  .[order(.[,"sample_id"], decreasing = TRUE),]
  
rownames(mmd) <- matrixrownames
### mmd <- data.frame(mmd[row_sorted,])
mmd <- data.frame(mmd)

# generate the barplot. this is generated as the annotation for the heatmap of the Patient_ID that is generated below.
### hm_dat <- hm_dat[row_sorted,]
#hm_dat <- hm_dat / hm_dat %>% rowSums()

# bring cell types in order (column order)
# col_order <- c("basophils", "eosinophils", "monocytes", "lymphocytes", "neutrophils")
# hm_dat <- hm_dat[, col_order]



colvec_patient_id <- structure(brewer.paired(mmd$patient_id %>% unique() %>% length()), names = mmd$patient_id %>% unique())
colvec_fill <- structure(unname(cols25(n = ncol(hm_dat))), names = colnames(hm_dat))
colvec_fill <- structure(met.brewer("Signac", n = ncol(hm_dat)), names = colnames(hm_dat))

# create annotation with cell type proportions
ha <- rowAnnotation(`Patient ID` = mmd$patient_id,
                    `Cell Type Proportion` =
                      anno_barplot(hm_dat, 
                                   gp=gpar(fill = colvec_fill),
                                   bar_width = 1,
                                   height = unit(25,"cm"),
                                   width = unit(11,"cm"),
                                   show_row_names = FALSE),
                        col = list(`Patient ID` = colvec_patient_id),
                    show_legend = FALSE)


dend <- as.dendrogram(hc)
#dend <- color_branches(dend, k = 4, col = c("gray50", "blue", "green", "red"), groupLabels = TRUE) # `color_branches()` returns a dendrogram object

# heatmap consisting of the patient_IDs. one color per patient
h1 <- Heatmap(mmd$timepoint, 
             col = colvec[2:5], 
             width = unit(0.5, "cm"), 
             #cluster_rows = dend,
             #row_dend_width = unit(1, "cm"),
             height = unit(25, "cm"),
             show_heatmap_legend = FALSE, 
             heatmap_legend_param = list(title = "Time point"),
             row_names_gp = gpar(cex=3),
             show_row_names = TRUE,
             right_annotation =  ha,
             column_labels = "Time point")

# plot the data
#pdf(paste0(path.out, "stacked_barplot_unclust.pdf"), height = 9, width = 16)
ht = grid.grabExpr(draw(h1))
grid.newpage()
pushViewport(viewport(angle = 270))
grid.draw(ht)
#dev.off()

#pdf(paste0(path.out, "stacked_barplot_legend_unclust.pdf"), height = 9, width = 16)

lgd1 <- Legend(labels = paste0("V", c(2,3,8,9)),
               title = "Time point",
               legend_gp = gpar(fill = colvec[2:5]),
               ncol = 1)


lgd2 <- Legend(labels = mmd$patient_id %>% unique(),
               title = "Patient ID",
               legend_gp = gpar(fill = brewer.paired(mmd$patient_id %>% unique() %>% length())),
               ncol = 1)

lgd3 <- Legend(labels = colnames(hm_dat),
               title = "Cell type",
               legend_gp = gpar(fill = met.brewer("Signac", n = ncol(hm_dat))),
               ncol = 1)

pd <- packLegend(lgd1, lgd2, lgd3, direction = "vertical", column_gap = unit(0.5, "cm"))

draw(pd)
#dev.off()
```


## Differential state analysis
Combining the expression data frame with the coldata gives arcsine transformed values. could take non-transformed values (i.e. "counts" assay) to get more familiar results.
```{r DS}
mfidf <- cbind(assay(sce, "exprs") %>% t() %>% as.data.frame(), colData(sce) %>% as.data.frame()) %>% dplyr::select(rownames(sce), "sample_id", "secondary")
mfidfga <- mfidf %>% 
  dplyr::group_by(sample_id, secondary) %>% 
  summarise_if(is.numeric, mean, na.rm = TRUE) %>%
  separate(sample_id, c("patient_id", "condition")) %>%
  gather(key = "marker", value = "mfi", rownames(sce))
mfidfga$condition <- mfidfga$condition %>% factor(levels = c("H", "V2", "V3", "V8", "V9"))

pdf(paste0(path.out, "7_Marker_Expression.pdf"), height = 9, width = 16, onefile = TRUE) #to save all the plots in one file
ggplot(mfidfga, aes(x = secondary, y = mfi, col = condition, fill = condition))+
    geom_boxplot(alpha = 0.5)+
    geom_point(position = position_dodge(width = 0.75))+
    #geom_line(aes(group = patient_id), col = "grey")+
    facet_wrap("marker", scales = "free_y")+
    theme_bw()+
    labs(col = "Time point", fill = "Time point", title = "Marker Expression on PBMCs", y = "MFI")+
    theme(axis.title.x = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank(),
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
    scale_color_manual(values = colvec)+
    scale_fill_manual(values = colvec)+
    #stat_compare_means(comparisons = list(c("MORNING", "EVENING")), method = "t.test", paired = TRUE)+
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))

for(mrkr in unique(mfidfga$marker)){
  p1 <- ggplot(mfidfga %>% dplyr::filter(marker == mrkr), aes(x = condition, y = mfi, col = condition, fill = condition))+
    geom_violin(alpha = 0.5, draw_quantiles = 0.5)+
    geom_point()+
    geom_line(aes(group = patient_id), col = "grey")+
    facet_wrap("secondary", scales = "free_y")+
    theme_bw()+
    labs(col = "Time point", fill = "Time point", title = paste0(mrkr, " MFI per cluster"), y = paste0("MFI (", mrkr, ")"))+
    theme(axis.title.x = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())+
    scale_color_manual(values = colvec)+
    scale_fill_manual(values = colvec)+
    stat_compare_means(comparisons = comparisons, method = "t.test", paired = TRUE)+ #CD4 Temra are not present in all samples, so the paired testing will fail there. still did the testing paired.
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))
    print(p1)
}
dev.off()
```


# CD4 T cells Abundance and stacked bar plot
```{r, fig.width=16}
scecd4 <- sce[,colData(sce)$primary == "CD4 T cells"]
colData(scecd4)$secondary <- colData(scecd4)$secondary %>% droplevels()


dtp <- colData(scecd4)[c("secondary", "sample_id")] %>% 
  table() %>% 
  as.data.frame() %>%
  mutate("copy" = sample_id) %>%
  separate(copy, c("patient_id", "condition")) %>%
  dplyr::rename("counts" = Freq)

totdtp <- colData(scecd4)[c("sample_id")] %>% table() %>% data.frame() %>% dplyr::rename("sample_id" = Var1, "totcounts" = Freq) #counts per sample to normalize counts by sample size. irrelevant in this case, since all samples have the same samplesize.
dtp <- merge(dtp, totdtp) %>% mutate("Freq" = counts/totcounts*100)


pdf(paste0(path.out, "6.2_Cluster_Abundance_of_CD4.pdf"), height = 9, width = 16)

comparisons <- list(c("V2", "V3"), c("V8", "V9"), c("V2", "V9"))
ggplot(dtp, aes(x = condition, y = Freq, col = condition))+
  geom_violin(aes(fill = condition), alpha = 0.5, draw_quantiles = 0.5)+
  #geom_jitter(width = 0.1)+
  geom_point()+
  geom_line(aes(group = patient_id), col = "grey")+
  facet_wrap("secondary", scales = "free_y")+
  theme_bw()+
  labs(col = "Time point", fill = "Time point", title = "Cluster Abundance per Time point", y = "Frequency of PBMCs (%)")+
  theme(axis.title.x = element_blank(),
        plot.title = element_text(hjust = 0.5),
        strip.background = element_blank())+
  scale_color_manual(values = colvec)+
  scale_fill_manual(values = colvec)+
  stat_compare_means(comparisons = comparisons, method = "t.test", paired = TRUE)+
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))


dev.off()

cur_df <- dtp %>%
  dplyr::filter(condition != "H") %>%
  dplyr::select(sample_id, secondary, Freq) %>%
  spread(key = secondary, value = Freq)

matrixrownames <- cur_df$sample_id 

# now we create a matrix from the data and cluster the data based on the cell fractions
hm_dat = as.matrix(cur_df[,-1])
rownames(hm_dat) <- as.character(matrixrownames)

# calculate distance and then cluster samples on cluster fraction
dd <- dist((hm_dat))
hc <- hclust(dd, method = "ward.D2")
row_sorted <- hc$labels

# now we generate the clinical metadata and order it in the same way as the celltype data
mmd <- cur_df %>% 
  dplyr::select(sample_id) %>%
  dplyr::mutate("copy" = sample_id) %>%
  separate(copy, into = c("patient_id", "timepoint"), sep = "_") %>%
  as.matrix()
  
rownames(mmd) <- matrixrownames
mmd <- data.frame(mmd[row_sorted,])

# generate the barplot. this is generated as the annotation for the heatmap of the Patient_ID that is generated below.
hm_dat <- hm_dat[row_sorted,]
#hm_dat <- hm_dat / hm_dat %>% rowSums()

# bring cell types in order (column order)
# col_order <- c("basophils", "eosinophils", "monocytes", "lymphocytes", "neutrophils")
# hm_dat <- hm_dat[, col_order]



colvec_patient_id <- structure(brewer.paired(mmd$patient_id %>% unique() %>% length()), names = mmd$patient_id %>% unique())
colvec_fill <- structure(unname(cols25(n = ncol(hm_dat))), names = colnames(hm_dat))
colvec_fill <- structure(met.brewer("Signac", n = ncol(hm_dat)), names = colnames(hm_dat))

# create annotation with cell type proportions
ha <- rowAnnotation(`Patient ID` = mmd$patient_id,
                    `Cell Type Proportion` =
                      anno_barplot(hm_dat, 
                                   gp=gpar(fill = colvec_fill),
                                   bar_width = 1,
                                   height = unit(25,"cm"),
                                   width = unit(11,"cm"),
                                   show_row_names = FALSE),
                        col = list(`Patient ID` = colvec_patient_id),
                    show_legend = FALSE)


dend <- as.dendrogram(hc)
#dend <- color_branches(dend, k = 4, col = c("gray50", "blue", "green", "red"), groupLabels = TRUE) # `color_branches()` returns a dendrogram object

# heatmap consisting of the patient_IDs. one color per patient
h1 <- Heatmap(mmd$timepoint, 
             col = colvec[2:5], 
             width = unit(0.5, "cm"), 
             cluster_rows = dend,
             row_dend_width = unit(1, "cm"),
             height = unit(25, "cm"),
             show_heatmap_legend = FALSE, 
             heatmap_legend_param = list(title = "Time point"),
             row_names_gp = gpar(cex=3),
             show_row_names = TRUE,
             right_annotation =  ha,
             column_labels = "Time point")

# plot the data
pdf(paste0(path.out, "stacked_barplot_CD4.pdf"), height = 9, width = 16)
ht = grid.grabExpr(draw(h1))
grid.newpage()
pushViewport(viewport(angle = 270))
grid.draw(ht)
dev.off()
pdf(paste0(path.out, "stacked_barplot_CD4_legend.pdf"), height = 9, width = 16)

lgd1 <- Legend(labels = paste0("V", c(2,3,8,9)),
               title = "Time point",
               legend_gp = gpar(fill = colvec[2:5]),
               ncol = 1)


lgd2 <- Legend(labels = mmd$patient_id %>% unique(),
               title = "Patient ID",
               legend_gp = gpar(fill = brewer.paired(mmd$patient_id %>% unique() %>% length())),
               ncol = 1)

lgd3 <- Legend(labels = colnames(hm_dat),
               title = "Cell type",
               legend_gp = gpar(fill = met.brewer("Signac", n = ncol(hm_dat))),
               ncol = 1)

pd <- packLegend(lgd1, lgd2, lgd3, direction = "vertical", column_gap = unit(0.5, "cm"))

draw(pd)
dev.off()
```

# Save SCE object
```{r save RDS}
saveRDS(object = sce, file = paste0(path.files, "SCE_downsammpled_subclustered_10k_2xflowsom.rds"))
```

# Session Info
```{r}
sessionInfo()
```

