---
title: "Flow_CATALYST"
author: "docaspar"
date: "12/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.width = 16, fig.height = 9, warning = FALSE, root.dir = ".")
```



# Libraries
```{r libs}
library(tidyverse)
library(SingleCellExperiment)
library(CATALYST)
library(flowCore)
library(xlsx)
library(readxl)
library(clustree)
library(diffcyt)
library(Rphenograph)
library(edgeR)
library(MetBrewer)
library(ggpubr)
library(pals)
library(ComplexHeatmap)
library(rasterpdf)
library(cowplot)
```


# Set paths
```{r set paths, warning=FALSE}
rm(list = ls())

path.data <- "../Data/20220505_Live"
path.files <- "../Files/"
path.out <- "../R_out/Unsupervised_10k_final/"

colvec <- met.brewer("Monet",n=20)[c(4,10,13,19,16)]

source("flow_functions.R")

clustvec <- brewer.paired(12)[c(3,1,7,4,10,2,6,9,8,5,11,12)]
names(clustvec) <- paste(1:12)
```

# Treg subclustering

Continuing with clustered rather than gated Tregs

## Filter Tregs, change type markers and re-run UMAP

To cluster Tregs based on the total clustering rather than the gated cells, we filter the secondary cluster ids for Tregs. We then change the markers (same as for clustering on manually gated Tregs) and re-create the UMAP.

```{r filter Tregs, fig.height=5, fig.width=8}
sce <- readRDS(file = paste0(path.files, "SCE_downsammpled_subclustered_10k_2xflowsom.rds"))
scetregs <- sce[,colData(sce)$secondary == "Treg"]

#treg_markers <- c("CCR7", "CD38", "CD31", "CD45RA", "PD1", "Foxp3", "HLADR", "CD25", "CD127", "TOX") #used this set of markers earlier, moved to the ones in the next line to make it more comparable with scRNAseq
treg_markers <- c("CCR7", "CD38", "CD31", "CD45RA", "PD1", "HLADR", "CD25", "CD127", "ICOS")
rowData(scetregs)[rowData(scetregs)$marker_name %in% treg_markers, "marker_class"] <- "type"
rowData(scetregs)[!rowData(scetregs)$marker_name %in% treg_markers, "marker_class"] <- "state"

type_markers(scetregs)
state_markers(scetregs)

set.seed(1212) #original seed was 1234
scetregs <- runDR(scetregs, dr = "UMAP", features = "type", cells = 500)

#pdf(paste0(path.out, "08_Treg_exploration.pdf"), onefile = TRUE, height = 5, width = 8)
plotDR(scetregs, "UMAP", color_by = "patient_id")
plotDR(scetregs, "UMAP", color_by = c(type_markers(scetregs), "Foxp3", "Ki67"), ncol = round(sqrt(16*length(type_markers(scetregs))/9)), assay = "exprs")

plot_density(scetregs, bins = 10, xlim = c(-10,8), ylim = c(-6.1,5)) + 
  scale_color_manual(values = colvec)
#dev.off()
```


## Phenograph clustering
I tried phenograph clusterings with k = 20, 30, 50, 80, 100 and 200. 100 seems to be ideal.
```{r Treg phenograph}
scetregs <- phenograph(scetregs, k = 100, scale = FALSE)

# set.seed(1234)
# phenoresults <- Rphenograph(t(assay(filterSCE(scetregs, marker_class == "type"), "exprs")), k = 100)
# 
# #Add phenograph cluster_ids to scetregs object
# colData(scetregs)$pheno <- as.factor(membership(phenoresults[[2]]))
# scetregs$cluster_id <- as.factor(membership(phenoresults[[2]]))
# metadata(scetregs)$cluster_codes <- data.frame(
#     custom = factor(levels(as.factor(membership(phenoresults[[2]]))), levels = levels(as.factor(membership(phenoresults[[2]])))))

rowData(scetregs)$marker_name
pdf(paste0(path.out, "09_Treg_subclustering_heatmap.pdf"), onefile = TRUE, height = 5, width = 8)
plotExprHeatmap(scetregs, features = c("CCR7", "CD45RA", "CD31", "CD25", "ICOS", "CD38", "HLADR",  "PD1", "CCR4", "CCR10"),
    by = "cluster_id", k = "custom",
    scale = "first", q = 0.01, perc = TRUE, col_dend = FALSE, bars = TRUE, col_clust = FALSE)
dev.off()
```


```{r Treg phenograph}
pdf(paste0(path.out, "09_Treg_subclustering.pdf"), onefile = TRUE, height = 5, width = 8)
plotExprHeatmap(scetregs, features = "type",
    by = "cluster_id", k = "custom",
    scale = "first", q = 0.01, perc = TRUE, col_dend = FALSE, bars = TRUE, col_clust = FALSE)

plotDR(scetregs, "UMAP", color_by = "pheno") + scale_color_manual(values = brewer.paired(12)[c(3,2,7,4:6,1,8:12)])
#plotDR(scetregs, "UMAP", color_by = c(type_markers(scetregs), "Foxp3", "Ki67"), ncol = round(sqrt(16*length(type_markers(scetregs)+2)/9)), assay = "exprs")
plotDR(scetregs, "UMAP", color_by = c(type_markers(scetregs), "Foxp3", "Ki67"), ncol = 4, assay = "exprs")+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

plot_density(scetregs, bins = 10, xlim = c(-10,8), ylim = c(-6.1,5)) + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        axis.title = element_blank()) + 
  scale_color_manual(values = colvec)

plot_density(scetregs, bins = 13, xlim = c(-10,8), ylim = c(-6.1,5)) + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        axis.title = element_blank()) +   
  scale_color_manual(values = colvec)
```


```{r Treg phenograph, fig.height=5, fig.width=15}
pdf(paste0(path.out, "density_facet.pdf"), height = 5, width = 15, onefile = TRUE)
plot_density(filterSCE(scetregs, scetregs$condition %in% c("H", "V2", "V3")), bins = 10, xlim = c(-10,8), ylim = c(-6.1,5)) + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        axis.title = element_blank()) +  
  scale_color_manual(values = colvec)

plot_density(filterSCE(scetregs, scetregs$condition %in% c("H", "V2", "V3")), bins = 13, xlim = c(-10,8), ylim = c(-6.1,5)) + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        axis.title = element_blank()) +  
  scale_color_manual(values = colvec)
dev.off()
```


```{r Treg phenograph, fig.height=5, fig.width=6}
pdf(paste0(path.out, "density_nofacet.pdf"), height = 5, width = 6, onefile = TRUE)
plot_density(filterSCE(scetregs, scetregs$condition %in% c("H", "V2", "V3")), bins = 10, xlim = c(-10,8), ylim = c(-6.1,5), facet = FALSE) + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        axis.title = element_blank()) +  
  scale_color_manual(values = colvec)

plot_density(filterSCE(scetregs, scetregs$condition %in% c("H", "V2", "V3")), bins = 13, xlim = c(-10,8), ylim = c(-6.1,5), facet = FALSE) + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        axis.title = element_blank()) +  
  scale_color_manual(values = colvec)

plot_density(filterSCE(scetregs, scetregs$condition %in% c("H", "V2", "V3")), bins = 13, xlim = c(-10,8), ylim = c(-6.1,5), facet = FALSE) + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        axis.title = element_blank()) +  
  scale_color_manual(values = c("darkgreen", "blue", "red"),)

plot_density(filterSCE(scetregs, scetregs$condition %in% c("V2", "V3")), bins = 13, xlim = c(-10,8), ylim = c(-6.1,5), facet = FALSE) + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        axis.title = element_blank()) +  
  scale_color_manual(values = c("blue", "red"),)

plot_density(filterSCE(scetregs, scetregs$condition %in% c("H", "V2", "V3")), bins = 13, xlim = c(-10,8), ylim = c(-6.1,5), facet = FALSE) + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        axis.title = element_blank()) +  
  scale_color_manual(values = c("darkgreen", "firebrick3", "tomato4"),)

plot_density(filterSCE(scetregs, scetregs$condition %in% c("V2", "V3")), bins = 13, xlim = c(-10,8), ylim = c(-6.1,5), facet = FALSE) + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        axis.title = element_blank()) +  
  scale_color_manual(values = c("firebrick3", "tomato4"),)
dev.off()
```

```{r, fig.height=5, fig.width=6}
pdf(paste0(path.out, "dotplot_nofacet.pdf"), height = 5, width = 6, onefile = TRUE)
plotDR(filterSCE(scetregs, scetregs$condition %in% c("H", "V2", "V3")), "UMAP", color_by = "condition")+
  scale_color_manual(values = c("darkgreen", "blue", "red"))+
  theme_cowplot()

plotDR(filterSCE(scetregs, scetregs$condition %in% c("V2", "V3")), "UMAP", color_by = "condition")+
  scale_color_manual(values = c("blue", "red"))+
  theme_cowplot()
dev.off()
```


```{r, fig.height=5, fig.width=15}
pdf(paste0(path.out, "dotplot_facet.pdf"), height = 5, width = 15, onefile = TRUE)
plotDR(filterSCE(scetregs, scetregs$condition %in% c("H", "V2", "V3")), "UMAP", color_by = "condition", facet_by = "condition")+
  scale_color_manual(values = c("darkgreen", "blue", "red"))+
  theme_cowplot()+
  theme(strip.background = element_blank())
dev.off()
```


```{r, fig.height=5, fig.width=10}
pdf(paste0(path.out, "dotplot_facet_woH.pdf"), height = 5, width = 10, onefile = TRUE)
plotDR(filterSCE(scetregs, scetregs$condition %in% c("V2", "V3")), "UMAP", color_by = "condition", facet_by = "condition")+
  scale_color_manual(values = c("blue", "red"))+
  theme_cowplot()+
  theme(strip.background = element_blank())
dev.off()
```

```{r Treg phenograph}
#raster_pdf(filename = paste0(path.out, "09b Subclustered_UMAP_combined_raster.pdf"), height = 9, width = 11, res = 200)
plotDR(scetregs, "UMAP", color_by = "pheno") +
  scale_color_manual(values = brewer.paired(12)[c(3,1,7,4,10,2,6,9,8,5,11,12)]) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
#dev.off()


markervec <- c("CD45RA", "CCR7", "CD31", "CD25", "CD127", "CD38", "HLADR", "PD1", "ICOS", "Ki67", "Foxp3")
#raster_pdf(filename = paste0(path.out, "09c Subclustered_UMAP_combined_raster.pdf"), height = 9, width = 14, res = 200)
plotDR(scetregs, "UMAP", color_by = markervec, ncol = 4, assay = "exprs")+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
#dev.off()

markervec <- c("CD45RA", "CCR7", "CD31", "CD25", "CD127", "CD38", "HLADR", "PD1", "ICOS", "Ki67", "Foxp3")
#raster_pdf(filename = paste0(path.out, "09c2 Subclustered_UMAP_combined_raster.pdf"), height = 9, width = 14, res = 200)
plotDR(scetregs, "UMAP", color_by = markervec, ncol = 6, assay = "exprs")+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
#dev.off()

#raster_pdf(filename = paste0(path.out, "09d Subclustered_UMAP_combined_raster.pdf"), height = 9, width = 14, res = 200)
plotDR(scetregs, "UMAP", color_by = c(type_markers(scetregs), "Foxp3", "Ki67"), ncol = 4, assay = "exprs")+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
#dev.off()

#raster_pdf(filename = paste0(path.out, "09e Subclustered_UMAP_combined_raster.pdf"), height = 9, width = 18, res = 200)
plotDR(scetregs, "UMAP", color_by = c("HLADR", "CD38"), ncol = 2, assay = "exprs")+
  theme_cowplot()+
  theme(strip.background = element_blank())
#dev.off()

#raster_pdf(filename = paste0(path.out, "09f Subclustered_UMAP_combined_raster.pdf"), height = 9, width = 11, res = 200)
plotDR(scetregs, "UMAP", color_by = "pheno") +
  scale_color_manual(values = brewer.paired(12)[c(3,1,7,4,10,2,6,9,8,5,11,12)]) +
  theme_cowplot()
#dev.off()

#pdf(paste0(path.out, "09h Subclustered_UMAP_combined_raster.pdf"), height = 5, width = 7)
plot_density(x = scetregs, color_by = "condition", colvec = colvec, xlim = c(-7.6, 10), ylim = c(-6, 6))+
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        strip.background = element_blank(),
        legend.position = "none")
#dev.off()
```

```{r}
act <- scetregs[,colData(scetregs)$pheno %in% c(7,9,12)]
a <- calculate_DS(act, cluster_column = "pheno", markers = rownames(act))

a
ggplot(a, aes(x = pheno, y = mfi))+
  geom_boxplot()+
  facet_wrap("marker", scales = "free_y")
```


```{r}
act <- scetregs[,colData(scetregs)$pheno %in% c(7,9,12)]
df <- cbind(assay(act, "exprs") %>% t(), colData(act)) %>% as.data.frame()
ggplot(df, aes(x = CD25, y = Ki67))+
  geom_point(aes(col = pheno), size = 0.5)+
  xlim(c(0, NA))+
  geom_smooth()+
  scale_color_manual(values = clustvec)
```



```{r}
scetregs <- runDR(scetregs, dr = "DiffusionMap", features = "type", cells = 500)
plotDR(scetregs, "DiffusionMap", color_by = c(type_markers(scetregs), "Foxp3", "Ki67"), ncol = round(sqrt(16*length(type_markers(scetregs))/9)), assay = "exprs")
plotDR(scetregs, "DiffusionMap", color_by = "pheno") + scale_color_manual(values = unname(cols25()))
```




```{r DA}
dtp <- colData(scetregs)[c("pheno", "condition", "patient_id")] %>% 
  table() %>% 
  as.data.frame() %>%
  mutate("sample_id" = paste(patient_id, condition, sep = "_")) %>%
  dplyr::rename("counts" = Freq)

totdtp <- colData(scetregs)[c("sample_id")] %>% table() %>% data.frame() %>% dplyr::rename("sample_id" = Var1, "totcounts" = Freq)
dtp <- merge(dtp, totdtp) %>% mutate("Freq" = counts/totcounts*100)
write.xlsx(dtp, file = paste0(path.out, "Treg_subcluster_abundance.xlsx"), row.names = FALSE)

pdf(paste0(path.out, "10_Treg_Cluster_Abundance.pdf"), height = 9, width = 16, onefile = TRUE) #to save all the plots in one file
comparisons <- list(c("V2", "V3"), c("V8", "V9"), c("V2", "V9"))
ggplot(dtp, aes(x = condition, y = Freq, col = condition))+
  geom_violin(aes(fill = condition), alpha = 0.5, draw_quantiles = 0.5)+
  #geom_jitter(width = 0.1)+
  geom_point()+
  geom_line(aes(group = patient_id), col = "grey")+
  facet_wrap("pheno", scales = "free_y")+
  theme_bw()+
  labs(col = "Time point", fill = "Time point", title = "Cluster Abundance per Time point", y = "Frequency (%)")+
  theme(axis.title.x = element_blank(),
        plot.title = element_text(hjust = 0.5),
        strip.background = element_blank())+
  scale_color_manual(values = colvec)+
  scale_fill_manual(values = colvec)+
  stat_compare_means(comparisons = comparisons, method = "t.test", paired = TRUE)+
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))

# plotAbundances(scetregs, k = "custom", group_by = "condition")
# plotAbundances(filterSCE(scetregs, condition != "H"), k = "custom", group_by = "patient_id")
dev.off()
```




## Differential state analysis
Combining the expression data frame with the coldata gives arcsine transformed values. could take non-transformed values (i.e. "counts" assay) to get more familiar results.
```{r DS}
mfidf <- cbind(assay(scetregs, "exprs") %>% t() %>% as.data.frame(), 
               colData(scetregs) %>% as.data.frame()) %>% 
  dplyr::select(rownames(scetregs), "sample_id", "pheno")

mfidfga <- mfidf %>% 
  dplyr::group_by(sample_id, pheno) %>% 
  summarise_if(is.numeric, mean, na.rm = TRUE) %>%
  separate(sample_id, c("patient_id", "condition")) %>%
  gather(key = "marker", value = "mfi", rownames(scetregs))
mfidfga$condition <- mfidfga$condition %>% factor(levels = c("H", "V2", "V3", "V8", "V9"))

pdf(paste0(path.out, "11_Treg_Marker_Expression.pdf"), height = 9, width = 16, onefile = TRUE) #to save all the plots in one file
# ggplot(mfidfga, aes(x = pheno, y = mfi, col = condition, fill = condition))+
#     geom_boxplot(alpha = 0.5)+
#     geom_point(position = position_dodge(width = 0.75))+
#     #geom_line(aes(group = patient_id), col = "grey")+
#     facet_wrap("marker", scales = "free_y")+
#     theme_bw()+
#     labs(col = "Time point", fill = "Time point", title = "Marker Expression on PBMCs", y = "MFI")+
#     theme(axis.title.x = element_blank(),
#           plot.title = element_text(hjust = 0.5),
#           strip.background = element_blank(),
#           axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
#     scale_color_manual(values = colvec)+
#     scale_fill_manual(values = colvec)+
#     #stat_compare_means(comparisons = list(c("MORNING", "EVENING")), method = "t.test", paired = TRUE)+
#     scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))

# for(mrkr in unique(mfidfga$marker)){
#   p1 <- ggplot(mfidfga %>% dplyr::filter(marker == mrkr), aes(x = condition, y = mfi, col = condition, fill = condition))+
#     geom_violin(alpha = 0.5, draw_quantiles = 0.5)+
#     geom_point()+
#     geom_line(aes(group = patient_id), col = "grey")+
#     facet_wrap("pheno", scales = "free_y")+
#     theme_bw()+
#     labs(col = "Time point", fill = "Time point", title = paste0(mrkr, " MFI per cluster"), y = paste0("MFI (", mrkr, ")"))+
#     theme(axis.title.x = element_blank(),
#           plot.title = element_text(hjust = 0.5),
#           strip.background = element_blank())+
#     scale_color_manual(values = colvec)+
#     scale_fill_manual(values = colvec)+
#     stat_compare_means(comparisons = comparisons, method = "t.test", paired = TRUE)+ #CD4 Temra are not present in all samples, so the paired testing will fail there. still did the testing paired.
#     scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))
#     print(p1)
# }
for(ph in unique(mfidfga$pheno)){
  p1 <- ggplot(mfidfga %>% dplyr::filter(pheno == ph), aes(x = condition, y = mfi, col = condition, fill = condition))+
    geom_violin(alpha = 0.5, draw_quantiles = 0.5)+
    geom_point()+
    geom_line(aes(group = patient_id), col = "grey")+
    facet_wrap("marker", scales = "free_y")+
    theme_bw()+
    labs(col = "Time point", fill = "Time point", title = paste0("Phenograph cluster ", ph, " MFI"), y = "MFI")+
    theme(axis.title.x = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())+
    scale_color_manual(values = colvec)+
    scale_fill_manual(values = colvec)+
    stat_compare_means(comparisons = comparisons, method = "t.test", paired = TRUE)+ #CD4 Temra are not present in all samples, so the paired testing will fail there. still did the testing paired.
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))
    print(p1)
}
dev.off()
```


```{r, eval=FALSE}
(merging_table <- read_excel(paste0(path.files, "merged_clusters_tregs2.xlsx")))
                            
scetregs <- mergeClusters(scetregs, k = "custom", table = merging_table, id = "tregs", overwrite = TRUE)

code <- cluster_codes(scetregs)[,c("custom", "tregs")]
coldata <- as.data.frame(colData(scetregs))
coldata <- merge(coldata, code, by.x = "pheno", by.y = "custom")
colData(scetregs) <- DataFrame(coldata[order(coldata$cell_id),])

pdf(paste0(path.out, "12_Treg_merging.pdf"), onefile = TRUE, height = 9, width = 16)
plotDR(scetregs, "UMAP", color_by = "tregs") +scale_color_brewer(palette = "Paired")
plotDR(scetregs, "UMAP", color_by = "pheno") + scale_color_manual(values = unname(cols25()))
plotDR(scetregs, "UMAP", color_by = type_markers(scetregs), ncol = round(sqrt(16*length(type_markers(scetregs))/9)))
plotExprHeatmap(scetregs, features = "type",
    by = "cluster_id", k = "tregs",
    scale = "first", q = 0.01, perc = TRUE, col_dend = FALSE, bars = TRUE, col_clust = FALSE)
dev.off()
```


```{r DA, eval=FALSE}
dtp <- colData(scetregs)[c("tregs", "condition", "patient_id")] %>% 
  table() %>% 
  as.data.frame() %>%
  mutate("sample_id" = paste(patient_id, condition, sep = "_")) %>%
  dplyr::rename("counts" = Freq)

totdtp <- colData(scetregs)[c("sample_id")] %>% table() %>% data.frame() %>% dplyr::rename("sample_id" = Var1, "totcounts" = Freq)
dtp <- merge(dtp, totdtp) %>% mutate("Freq" = counts/totcounts*100)

pdf(paste0(path.out, "13_Treg_merged_Cluster_Abundance.pdf"), height = 5, width = 8, onefile = TRUE) #to save all the plots in one file
comparisons <- list(c("V2", "V3"), c("V8", "V9"), c("V2", "V9"))
ggplot(dtp, aes(x = condition, y = Freq, col = condition))+
  geom_violin(aes(fill = condition), alpha = 0.5, draw_quantiles = 0.5)+
  #geom_jitter(width = 0.1)+
  geom_point()+
  geom_line(aes(group = patient_id), col = "grey")+
  facet_wrap("tregs", scales = "free_y")+
  theme_bw()+
  labs(col = "Time point", fill = "Time point", title = "Cluster Abundance per Time point", y = "Frequency (%)")+
  theme(axis.title.x = element_blank(),
        plot.title = element_text(hjust = 0.5),
        strip.background = element_blank())+
  scale_color_manual(values = colvec)+
  scale_fill_manual(values = colvec)+
  stat_compare_means(comparisons = comparisons, method = "t.test", paired = TRUE)+
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))

plotAbundances(scetregs, k = "tregs", group_by = "condition")
plotAbundances(filterSCE(scetregs, condition != "H"), k = "tregs", group_by = "patient_id")
dev.off()
```



## Differential state analysis
Combining the expression data frame with the coldata gives arcsine transformed values. could take non-transformed values (i.e. "counts" assay) to get more familiar results.
```{r DS, eval=FALSE}
mfidf <- cbind(assay(scetregs, "exprs") %>% t() %>% as.data.frame(), colData(scetregs) %>% as.data.frame()) %>% dplyr::select(rownames(scetregs), "sample_id", "tregs")
mfidfga <- mfidf %>% 
  dplyr::group_by(sample_id, tregs) %>% 
  summarise_if(is.numeric, mean, na.rm = TRUE) %>%
  separate(sample_id, c("patient_id", "condition")) %>%
  gather(key = "marker", value = "mfi", rownames(scetregs))
mfidfga$condition <- mfidfga$condition %>% factor(levels = c("H", "V2", "V3", "V8", "V9"))

pdf(paste0(path.out, "14_Treg_merged_marker_expression.pdf"), height = 9, width = 16, onefile = TRUE) #to save all the plots in one file
# ggplot(mfidfga, aes(x = tregs, y = mfi, col = condition, fill = condition))+
#     geom_boxplot(alpha = 0.5)+
#     geom_point(position = position_dodge(width = 0.75))+
#     #geom_line(aes(group = patient_id), col = "grey")+
#     facet_wrap("marker", scales = "free_y")+
#     theme_bw()+
#     labs(col = "Time point", fill = "Time point", title = "Marker Expression on PBMCs", y = "MFI")+
#     theme(axis.title.x = element_blank(),
#           plot.title = element_text(hjust = 0.5),
#           strip.background = element_blank(),
#           axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
#     scale_color_manual(values = colvec)+
#     scale_fill_manual(values = colvec)+
#     #stat_compare_means(comparisons = list(c("MORNING", "EVENING")), method = "t.test", paired = TRUE)+
#     scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))

# for(mrkr in unique(mfidfga$marker)){
#   p1 <- ggplot(mfidfga %>% dplyr::filter(marker == mrkr), aes(x = condition, y = mfi, col = condition, fill = condition))+
#     geom_violin(alpha = 0.5, draw_quantiles = 0.5)+
#     geom_point()+
#     geom_line(aes(group = patient_id), col = "grey")+
#     facet_wrap("tregs", scales = "free_y")+
#     theme_bw()+
#     labs(col = "Time point", fill = "Time point", title = paste0(mrkr, " MFI per cluster"), y = paste0("MFI (", mrkr, ")"))+
#     theme(axis.title.x = element_blank(),
#           plot.title = element_text(hjust = 0.5),
#           strip.background = element_blank())+
#     scale_color_manual(values = colvec)+
#     scale_fill_manual(values = colvec)+
#     stat_compare_means(comparisons = comparisons, method = "t.test", paired = TRUE)+ #CD4 Temra are not present in all samples, so the paired testing will fail there. still did the testing paired.
#     scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))
#     print(p1)
# }
for(ph in unique(mfidfga$tregs)){
  p1 <- ggplot(mfidfga %>% dplyr::filter(tregs == ph), aes(x = condition, y = mfi, col = condition, fill = condition))+
    geom_violin(alpha = 0.5, draw_quantiles = 0.5)+
    geom_point()+
    geom_line(aes(group = patient_id), col = "grey")+
    facet_wrap("marker", scales = "free_y")+
    theme_bw()+
    labs(col = "Time point", fill = "Time point", title = paste0("Phenograph cluster ", ph, " MFI"), y = "MFI")+
    theme(axis.title.x = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())+
    scale_color_manual(values = colvec)+
    scale_fill_manual(values = colvec)+
    stat_compare_means(comparisons = comparisons, method = "t.test", paired = TRUE)+
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))
    print(p1)
}
dev.off()
```

```{r}
coldata.sce <- colData(sce)
coldata.tregs <- colData(scetregs)
coldata <- merge(coldata.sce, coldata.tregs[,c("cell_id", "pheno")], by = "cell_id", all = TRUE)
coldata$secondary <- as.character(coldata$secondary)
coldata$pheno <- as.character(coldata$pheno)
coldata[is.na(coldata$pheno), "pheno"] <- coldata[is.na(coldata$pheno), "secondary"]
coldata$pheno <- factor(coldata$pheno, c(paste(1:12), "Tcon", "CD8 T cells", "DN T cells", "DP T cells", "Dendritic cells", "classical Monocytes", "intermediate Monocytes", "non-classical Monocytes", "CD19 B cells", "CD56bright NK cells", "CD56dim NK cells", "other PBMCs"))

colData(sce) <- DataFrame(coldata[order(coldata$cell_id),])

pdf(paste0(path.out, "15_Treg_Backclustering.pdf"), height = 9, width = 16, onefile = TRUE) #to save all the plots in one file
plotDR(sce, "UMAP", color_by = "pheno") +scale_color_manual(values = unname(cols25()))
plotDR(sce, "UMAP", color_by = type_markers(sce), ncol = round(sqrt(16*length(type_markers(sce))/9)))
#dev.off()

scecd4 <- sce[,colData(sce)$primary == "CD4 T cells"]

cd4_markers <- c("CD127", "CD25", "Foxp3")
rowData(scecd4)[rowData(scecd4)$marker_name %in% cd4_markers, "marker_class"] <- "type"
rowData(scecd4)[!rowData(scecd4)$marker_name %in% cd4_markers, "marker_class"] <- "state"

type_markers(scecd4)

set.seed(1234)
scecd4 <- runDR(scecd4, dr = "UMAP", features = "type", cells = 500)
plotDR(scecd4, "UMAP", color_by = c(type_markers(scetregs), "Foxp3", "Ki67"), ncol = 6, assay = "exprs")
plotDR(scecd4, "UMAP", color_by = "pheno", assay = "exprs")+scale_color_manual(values = unname(cols25()))
dev.off()
```

```{r}
flowsom(x = scetregs, markers = type_markers(scetregs), )






```

