---
title: "01_Protein_read_data"
author: "toobiwankenobi"
date: "2020-08-04"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

## Introduction
This file will load the single-cell data and store it in an SingleCellExperiment data container. In order to successfully run this script, several .csv files have to provided in the data folder of this repository.

# Preparations

## Load libraries

First, we will load the libraries needed for this part of the analysis.

```{r load-libraries, message=FALSE}
library(tidyverse)
library(data.table)
library(SingleCellExperiment)
library(rstudioapi)
library(xlsx)
library(CATALYST)
```

## Read the data

```{r load data}
#Clear R's memory
rm(list=ls())

#set wd to location of this .Rmd worksheet. ADAPT PATH such that it points from .Rmd location to cpout folder!
setwd(dirname(getActiveDocumentContext()$path)) 
path.data = "../Data/cpout_3pix2/"
path.files <- "../R_Files/cpout_3pix2/"
path.out <- "../R_out/cpout_3pix2/"

# load the single cell data
cells <- as.data.frame(fread(file = paste0(path.data, "cell.csv"),stringsAsFactors = FALSE))

# load the image level metadata
image_mat <- as.data.frame(read.csv(file = paste0(path.data, "Image.csv"),stringsAsFactors = FALSE))

# load the panel information
###THIS NEEDS FIXING###
panel_mat <- read.csv(file = paste0(path.data, "panel.csv"), sep= ",",  stringsAsFactors = FALSE )
#panel_mat <- read.csv(file = paste0(path.data, "example_panel_marker_classes.csv"), sep= ";",  stringsAsFactors = FALSE )

# get an example file that contains the channel order
tags <- read.csv(file = paste0(path.data, "full_channelmeta.csv"), header = FALSE)

# load acquisition meta data
acquisition_meta <- read.csv(file = paste0(path.data, "acquisition_metadata.csv"), stringsAsFactors = FALSE)

# load the clinical data. this is a clinical datatable that already contains the ImageNumbers. It has been prepared in the clinical metadata preparation.Rmd script (prepared for RNA and protein dataset separately)
#clinical_mat <- read.csv("data/data_for_analysis/protein/clinical_data_protein.csv",stringsAsFactors = FALSE)
```

# Pre-processing

## Generate the counts data frame

```{r get count data }
cur_counts <- cells[,grepl("Intensity_MeanIntensity_FullStackFiltered",colnames(cells))]
```

## Get the scaling factor
the single cell data needs to be multiplied with the scaling factor (16 bit)

```{r scale the count data}
cur_counts <- cur_counts * image_mat$Scaling_FullStack[1]

# to order the channels according to channel number
channelNumber <- as.numeric(sub("^.*_c", "", colnames(cur_counts)))

cur_counts <- cur_counts[,order(channelNumber,decreasing = FALSE)]
```

## Prepare the cell-level metadata
this data frame contains the metadata for ever single cell and will later on be the colData in the single cell experiment object

the metadata will also have an entry called "Parent_nuclei" which holds information to the ObjectNumber of the Nuclei that the cell was derived from.
due to the down-scaling of the images some nuclei are lost and thus some cells do not have a Parent_nuclei
```{r column level metadata}
#docaspar: add timepoint information to cells. merge acquisition_meta to image_mat, then image_mat to cells
sub_acquisition_meta <- acquisition_meta %>% dplyr::select(start_timestamp, description)
"Metadata_start_timestamp" %in% colnames(image_mat)
image_mat <- merge(image_mat, sub_acquisition_meta, by.x = "Metadata_start_timestamp", by.y = "start_timestamp")
image_mat_sub <- image_mat %>% dplyr::select(ImageNumber, description) %>% separate(description, c("patient_id", "timepoint", "ROI"))

cells <- merge(cells, image_mat_sub, by= "ImageNumber")

cells[!grepl("Intensity", colnames(cells))] %>% head()
cells$timepoint_factor <- factor(cells$timepoint, levels = c("Before", "After"))

cell_meta <- DataFrame(CellNumber = cells$ObjectNumber,
                       sample_id = as.character(cells$ImageNumber),
                       ImageNumber = cells$ImageNumber,
                       Center_X = cells$Location_Center_X,
                       Center_Y = cells$Location_Center_Y,
                       NumberOfNeighbors = cells$Neighbors_NumberOfNeighbors_8,
                       Timepoint = cells$timepoint_factor,
                       condition = cells$timepoint_factor,
                       Patient_ID = cells$patient_id)

# add a unique cellID to each cell consisting of "dataset"+"ImageNumber"+"ObjectNumber"
cell_meta$cellID <- paste("cell",cell_meta$ImageNumber,cell_meta$CellNumber, sep = "_")

rownames(cell_meta) <- cell_meta$cellID
```

## Prepare the row-level metadata (panel/marker information)
here we prepare all the metadata for the rows in the single cell experiment object (rowData)

```{r row level metadata}
# assign channel number, matching with the measured channles in cellprofiler
# the channel number should be double-checked with the channel order in the _full.csv file in the tiff folder to see if the order matches (e.g. Y89 can cause a mismatch)

# the channel numbers are the rownumbers in the "tags" file that we create above
tags$channel <- as.numeric(rownames(tags))
colnames(tags) <- c("Metal.Tag","channel")


# include the channel information in the panel metadata (panel_mat)
panel_mat <- merge(panel_mat,tags,by="Metal.Tag")

# now we order the panel metadata by channel. therefore we first modify the column names

panel_mat <- panel_mat[order(panel_mat$channel,decreasing = FALSE),]

# we also revise the nomenclature of the clean targets to not contain special characters like "-" etc
panel_mat$clean_target <- panel_mat$Target %>% str_replace_all("-", "")
panel_mat$marker_name <- panel_mat$Target %>% str_replace_all("-", "")

rownames(panel_mat) <- panel_mat$clean_target
```

# Create SCE object

## Create the single cell experiment object

```{r create single cell object}
# create the SCE object
sce <- SingleCellExperiment(assays = list(counts = t(cur_counts)))

# Set marker name as rownames and cellID as colnames
rownames(sce) <- rownames(panel_mat)
colnames(sce) <- rownames(cell_meta)

# add the column and row metadata
colData(sce) <- cell_meta
rowData(sce) <- panel_mat

ei <- data.frame(cell_meta) %>% dplyr::select(sample_id, condition, Patient_ID)
rownames(ei) <- NULL
metadata(sce)$experiment_info <- ei[!duplicated(ei$sample_id),]
ei(sce)
# asinh transformed counts as well as add the nuclear count data
assay(sce, "exprs") <- asinh(counts(sce))
rowData(sce)
```

## Assign the clinical data to the metadata slot

```{r}
#order according to ImageNumber
# clinical_mat <- clinical_mat[order(clinical_mat$ImageNumber),]
# metadata(sce) <- as.list(clinical_mat)
```

## Save the SCE object

```{r save sce object}
#saveRDS(sce,file = paste0(path, "../SCEobjects/sce_2_3pix2.rds"))
colData(sce) %>% data.frame
saveRDS(sce,file = paste0(path.files, "sce_3pix2.rds"))
```
